---
title: Regression with Time Series
---

```{r}
#| label: setup
#| output: false
#| code-fold: true

## load libraries
library(readxl)
library(tidyverse)
library(knitr)
library(lme4)
library(caret)
library(sjPlot)
library(nlme)
library(astsa)

###### Load (synthetic data) files:
synth_ww <- read.csv("Data/Synthetic_WW_PR.csv")

```

## A correlated error regression model

We saw that the linear model's independence assumption was violated due to temporal correlation present in the residuals. The below fits a time series model to the data which explicitly models this correlation using the `correlation` argument of the `gls` function in the `nlme` package.

```{r}
# fit model with AR(1) structure
m2 <- nlme::gls(SPR ~ SWWlog10cpd, data =synth_ww, correlation = nlme::corARMA(p=1, q = 0))
summary(m2)
```

## Has the temporal structure been captured?

```{r}
astsa::acf1(residuals(m2, type = "normalized"))
```

No-- the plot here still looks very similar to the one for linear regression, indicating the temporal structure has not been captured.

## Increasing lag

The previous model only accounted for correlation for one lag (one week in the past). What about looking further back?

```{r}
# fit model with AR(2) structure
 m3 <- nlme::gls(SPR ~ SWWlog10cpd, data =synth_ww, correlation = nlme::corARMA(value = c(0.8, 0.04), p=2, q = 0))
summary(m3)
```

## Has the temporal structure been captured?

```{r}
astsa::acf1(residuals(m3, type = "normalized"))
```
