---
title: "Interactive Map with Clickable Points Displaying Graphs"
author: "Julia Schedler & Jose Palacio"
format: html
editor: visual
---

Can we have an interactive map where you can click a WWTP and have a graph pop up? (so like clickable tabs, but instead of tabs, it's a map)

Starting point:

-   <https://tmieno2.github.io/R-as-GIS-for-Economists/interactive-view-of-an-sf-object.html>

-   your code for the mapping of a gif

Goals:

1.  minimal example– one clickable map that allows you to see the 37 time series plots
2.  Create tutorial for anything that isn't easy for an R user comfortable with sf, ggplot, tidyverse
3.  Create a demonstration of the weekly workflow (if minimal example is created using wwtp_070824.Rdata, then show how to update the minimal example using a new data set like wwtp_071524.Rdata)

Stretch goal (don't think about this now)

-   have a slider for time points, and clickable map

-   have the ability to filter with different viruses

## Jose's stuff goes below here!


Yes, we can! Here's a minimal example of how to create an interactive map with clickable points that display a graph when clicked. The example uses the `mapview` package to create the interactive map and the `popupGraph` function to display the graph when a point is clicked.

Use case: sending weekly reports of the weekly wwepi system to the health department

1.  Libraries needed

-   The `mapview` package (for interactive maps) has many dependencies that need to be loaded.
-   The `popupGraph` function is used to display the time series plot when a point is clicked.
-   The function `mapview` and `popupGraph` are functions from the `mapview` package.
-   Other libraries are used for data manipulation, visualization, and other tasks.
-   The `options(warn = -1)` line is used to suppress warnings.

```{r}
#| results: asis
#| warning: false
#| echo: true
#| message: false
#| code-fold: true
#| outout: false
#| label: MARSS-libraries

library(MARSS)
library(tidyverse)
library(KFAS)
library(ggplot2)
library(astsa)
library(dplyr)
library(tseries)
library(knitr)
library(latexpdf)
library(tibble)
library(kableExtra)
library(forecast)
library(qcc)
library(ggnewscale)
library(sf)
library(png)
library(rnaturalearth)
library(leaflet)
library(leafpop)
library(leaflet.providers)
library(mapview)
options(warn = -1)
```

2.  Load the data

```{r}
#| warning: false
#| echo: true
#| message: false
#| code-fold: true
#| label: load-data

load("Data/all_ts_observed_2023_05_29")
load("Data/case_wwtp_ag") 
source("Code/fplot.R")
```

3.  Load the shapefile:
    - Note that we had to make some modifications to the shapefile to match the data.
    - We had to rename the column to "Name" and remove some WWTPs that were not in the data.
    - We also had to put the WWTPs in alphabetical order for easier computation.
    - We also had to remove Sims Bayou and Kingwood from the shapefile.
    - Finally, we had to add an id column to the shapefile.
    - The shapefile is loaded using the `read_sf` function from the `sf` package.
    - The functions `mutate`, `filter`, and `arrange` are from the `dplyr` package.
    - the function `toupper` and `rbind` are from the `base` package.
    
```{r}
#| warning: false
#| echo: true
#| message: false
#| code-fold: true
#| label: shape-data

# Load the shapefile
wwtp_shp <- sf::read_sf("Data/WWTP_061621/WWTP_061621.shp")
names(wwtp_shp)[1] <- "Name" # rename column
wastelake_shp <- wwtp_shp %>% 
  dplyr::filter(Name %in% c("West Lake Houston")) %>% # filter
  dplyr::mutate(Name = "West Lake") # rename
wwtp_shp <- base::rbind(wwtp_shp, wastelake_shp) 
wwtp_shp <- wwtp_shp[wwtp_shp$Name != "West Lake Houston",] # remove west lake
wwtp_shp <- wwtp_shp[wwtp_shp$Name != "MC MUD #48",] # remove MC MUD #48
wwtp_shp$Name <- base::toupper(wwtp_shp$Name) # upper case
## put in alphabetical order for easier computation
wwtp_shp <- wwtp_shp %>% 
  dplyr::arrange(Name)
wwtp_shp$id <- 1:37 
# Omit Sims Bayou and Kingwood
wwtp_shp <- wwtp_shp %>% 
  dplyr::filter(Name != "SIMS BAYOU" & Name != "KINGWOOD")
```

4.  Time series plots
    - The time series plots are created using the `ggplot2` package.
    - The `group_nest` function is used to group the data by WWTP and create a list of data frames.
    - The `map` function is used to create a list of time series plots for each WWTP.
    - The plots are created using the `ggplot` function and the `geom_line` and `geom_point` functions.
    - The `scale_shape_manual` function is used to set the shape of the points in the plot.

```{r}
#| warning: false
#| echo: true
#| message: false
#| code-fold: true
#| label: time-series

# Time Series Plots
ts_plots <- all_ts_observed %>%
          dplyr::group_nest(WWTP, keep = T) %>% 
          tibble::deframe() %>%
          purrr::map(., ~ {
            ggplot2::ggplot(.x, aes(x = date, y = log10cpl)) +
              ggplot2::theme_minimal()+
              ggplot2::geom_line() + 
              ggplot2::labs(title = .x$WWTP, x= "Date", y = "Log10 Copies/L-WW", color = "") +
              ggplot2::geom_point(aes(x=date, y=log10cpl)) +
              ggplot2::labs(color = "") +
              ggplot2::scale_shape_manual(values = c(8, 19))
            })
```

5.  Create the interactive map with clickable points. Note that: 
    - The `popupGraph` function is used to display the time series plot when a 
      point is clicked. 
    - Also, we omit Sims Bayou, Sims Bayou North, Sims Bayou South, 
      Kingwood Central, and Kingwood West from the map. 
    - One time series data set (visualization) for each of the WWTPs

```{r}
#| warning: false
#| echo: true
#| message: false
#| code-fold: true
#| label: time-series-map

# Omit Sims Bayou, Sims Bayou North, Sims Bayou South, Kingwood Central, and Kingwood West
mapview::mapView(wwtp_shp, zcol = "Name", legend = FALSE, 
        popup = leafpop::popupGraph(ts_plots[-c(15, 16, 25, 26, 27)], width = 300, height = 300))
```

6.  Update the time series plots with new data points. Each week, a report is sent where each time series plot has one new observation
    -   This run by just reading in a new data file and updating the existing data with new observations.
    -   The new data is simulated for the current week by adding random noise to the existing data.
    -   The new data only contains the new observations for the current week at each WWTP.
    
```{r}
#| warning: false
#| echo: true
#| message: false
#| code-fold: true
#| label: Updated-time-series-plots

# Function to load the new data file
load_new_data <- function(file_path) {
  new_data <- utils::read.csv(file_path)
  return(new_data)
}

# Function to update the existing data with new observations
update_data <- function(existing_data, new_data) {
  updated_data <- base::rbind(existing_data, new_data)
  return(updated_data)
}

existing_data <- all_ts_observed

# Simulate new data for the current week
new_data <- existing_data %>%
  dplyr::filter(date == max(date)) %>%
  dplyr::mutate(date = date + 7,  # Simulate next week's data
         log10cpl = log10cpl + rnorm(n(), 0, 0.1))

# Update the data
all_ts_observed <- update_data(existing_data, new_data)

# Function to create time series plots with new data points highlighted in red
updated_ts_plot <- function(data, new_data) {
  ggplot2::ggplot(data, aes(x = date, y = log10cpl)) +
    ggplot2::theme_minimal() +
    ggplot2::geom_line() +
    ggplot2::labs(title = unique(data$WWTP), x = "Date", y = "Log10 Copies/L-WW", color = "") +
    ggplot2::geom_point(aes(x = date, y = log10cpl)) +
    ggplot2::geom_point(data = new_data, aes(x = date, y = log10cpl), color = "red", size = 3) +
    ggplot2::labs(color = "") +
    ggplot2::scale_shape_manual(values = c(8, 19))
}

# Generate the plots with highlighted new points
ts_plots <- all_ts_observed %>%
  dplyr::group_nest(WWTP, keep = TRUE) %>%
  tibble::deframe() %>%
  purrr::map(., ~ {
    new_data_point <- new_data %>% 
      dplyr::filter(WWTP == .x$WWTP[1])
    updated_ts_plot(.x, new_data_point)
  })

# Remove specific indices if necessary (15, 16, 25, 26, 27)
ts_plots <- ts_plots[-c(15, 16, 25, 26, 27)]
```

7.  Update the interactive map with the new time series plots.
    -   1 time series data set (visualization) for each of the wwtps

```{r}
#| warning: false
#| echo: true
#| message: false
#| code-fold: true
#| label: Updated-time-series-map

# Create the map with popup graphs
mapview::mapview(wwtp_shp, zcol = "Name", legend = FALSE, 
               popup = leafpop::popupGraph(ts_plots, width = 300, height = 300))

```
